Fast Sorting on Modern Computers

### 2005-06-14 docs/fyp_report/quick.tex
moved doc dirs into docs/ moved scripts into scripts/ moves counters/ into data/

%\section{Description}
\label{quick}
Quicksort is an in-place, unstable sort. It was first described in
\cite{HoarePartition61} and \cite{HoareQuicksort61}, and later in more detail in
\cite{Hoare62}. Sedgewick did his PhD on quicksort, and several important
improvements to quicksort are due to this. They are discussed in
\cite{Sedgewick78} and \cite{Sedgewick02}. \cite{BentleyMcIlroy93} discusses
considerations for a fast C implementation. \cite{SedgewickBentley02} discusses
equal keys and proves that quicksort has the lowest complexity of any comparison
based sort.

@@

Once the entire array has been put into containers, each container is emptied
in turn and the keys put back into the array. When a container is empty, the
pivot greater than that container is put into position, which is guaranteed to
be its final position.  The indices of these positions is pushed onto the
stack, and the subarray is then sorted.

While emptying the containers back into the array, an opportunity is taken to
find the smallest key in the leftmost partition. This is placed as a sentinel,
ensuring that every partition has a sentinel, and that no bounds checks are
required.

@@

The instruction count, shown in Figure \ref{quick instructions} is as
predicted. The expected increases in instruction count for multi-quicksort was
due to a test with a smaller data set and smaller constants. The actual results
were higher than expected: being 63\% and 40\% greater, for
sequential and binary search versions, respectively.
%TODO check

The cache results in Figure \ref{quick level 2 misses} were similar to
predictions, except that the multi-quicksorts did not behave as expected, and
performed worse than the memory-tuned version. Not only do they have more
misses per key, but the rate at which the misses increase with size is also
worse for the multi quicksorts. The tuned version performed better than the
base version, as predicted.

@@

The sequential search stops the increase of branch misses, since every key will
only take one miss to find its proper place. The results of this on cycle count,
shown in Figure \ref{quick cycles} are enough to make up for the increase in
instruction count, which was significantly reduced by using the binary search.
The tuned quicksort, however, had the lower running time and even base
quicksort out-ran the multi-quicksorts.

LaMarca's results are again similar to the results presented here, but not
identical. The shape of the instruction count graph is the same in both cases,
but that of the cache misses is not. The ratio of the cache misses between base
and cache quicksort are the same in both sets of results, but the number of
cache misses in multi-quicksort is much higher here than it is in LaMarca's
findings. In addition, the shape is different. LaMarca's multi-quicksort's
cache misses plateau, while the results here show an increasing slope.

A previous result, however, with a different implementation showed the same
shape, but different results to LaMarca's. That implementation waited until all
the partitioned keys were put back into the array, instead of being sorted
straight away. Those results plateaued once mutli-quicksort took effect.

\section{Predictor results}
\sixplot{base_quicksort-0.eps}{base_quicksort9-0.eps}{cache_quicksort-0.eps}{multi_quicksort-0.eps}{sequential_multi_quicksort-0.eps}{}{base, cache and multi quicksorts and base quicksort9}
\sixplot{base_quicksort1-0.eps}{base_quicksort-0.eps}{base_quicksort5-0.eps}{base_quicksort7-0.eps}{base_quicksort9-0.eps}{}{base quicksort without a median, with median of 3, and with pseudo median of 5, 7 and 9}

%TODO a bit to rewrite

Theres two important bits: comparison between the types of sorts, and comparisons between the different number of things tried as medians.

Firstly, cache quicksort and base quicksort only idffer in how they deal with
the insertion. Base quicksort and base quicksort9 have similar numbers of
insertion branches, and a very similar ratio (2/1 correct/incorrect) of
mispredictions. Comparing base quicksort to cahce quicksort (they both use
median of 3), there is actually a decrease in the number of branches handled by
insertion sort, instead of an increase. An increase is expected (thats why
sedgewick moves the insertion to the end), though when you take into account
the sentinel and the outer loop, there probably is an increase in instructions
at least. Base quicksort has 35\% miss rate, but cache has a 34\%. Not a big
difference, but the point is that there is a reduction in branches, but of the
reduced branches, 50\% were unpredictable. I dont think theres anything
interesting in this fact, however.

When going to multi-quicksort, there si quite a drop in the number fo branches.
Once the sort is larger than the cache, the number fo steps increases linearly
(since it does a sort of counting like radix sort). Theres a decrease of 19 \%.
Since the sort at this point is NlogM where M is the size of the cache (in
keys), there shouldnt be a change in misprediction rates. And there sint.
Before: 63\%. After, 63\%. So no difference.

Theres a lot of branches in the multi quicksort. WE saw in other results that
the mere instruction cost of this bogs down to algorithm significantly. in the
binary search version, there are more than twice as many mispredictions,
despite the fact that there are less than half the branches. (note that i dont
think this was a wonderful binary search, its heavily biased towards the
checking the left). binary search predicts correctly 57 percent of the time,
meaning that it mostly changes direction haphazardly during the search. No
wonder it performs so badly.

Note that sequential multi-quicksort has less mispredictions than cache quicksort.

Part 2: comparing different ways of randomly partitioning:

Ways of getting the partition:
1 - just take the rightmost element. Note that there are no sentinels, so you have to put a bounds check back in
3 - median of(0,1,2) halves
5 - median of(0, (median of(1,2,3), 4) quarters
7 - median of(median of(0,1,2), 3, median of(4,5,6)) sixths
9 - median of(median of(0,1,2), median of(3,4,5), median of(6,7,8)) eights

misprediction rates:
1 - 29.4\%
3 - 35.9\%
5 - 38.4\%
7 - 40.0\%
9 - 41.7\%

The misprediction rates point to the problem. They're just not being evenly divided. The number of times the partitions end is the same, roughly, each time. However, the number of actual branches is much higher in the case of quicksort1, and decreases all the way down. At the same time, the number of mispredictions creeps up. And hence the big change in the rate of mispredictions. As predicted, the better the median, the better the partition, and the less correct predictions pretending to be useful. There must be a 50\% rate waiting to get out, so a little more than median of 9 maybe. Or a lower cut-off point (it cuts out for partitions of less than 40 keys, and uses plain old median of 3).

the rate or number of insertion branches doesnt change at all.

%TODO up to here

\section{Future Work}
The threshold in quicksort is the point at which a simpler sort is used to sort
the partition. A larger threshold will results in fewer checks for a sentinel. A
smaller threshold means that the instruction count due to the insertion sort
will be lower, but more partition steps may be taken. A smaller threshold also
makes it more likely that the keys to be sorted all fit in a single cache
block. However, the partition is likely to occur across a cache block boundary,
though this is not necessarily a bad thing; the next partition to be sorted will
be adjacent to the current block, and will already be in memory. Observing how
results, especially instruction count and branch misses, vary with changes 
to the threshold may be interesting.

>>>>

%\section{Description}

Quicksort is an in-place, unstable sort. It was first described in
\cite{HoarePartition61} and \cite{HoareQuicksort61}, and later in more detail in
\cite{Hoare62}. Sedgewick did his PhD on quicksort, and several important
improvements to quicksort are due to this. They are discussed in
\cite{Sedgewick78} and \cite{Sedgewick02}. \cite{BentleyMcIlroy93} discusses
considerations for a fast C implementation. \cite{SedgewickBentley02} discusses
equal keys and proves that quicksort has the lowest complexity of any comparison
based sort.

@@

Once the entire array has been put into containers, each container is emptied in
turn and the keys put back into the array. When a container is empty, the pivot greater than that
container is put into position, which is guaranteed to be its final position.
The indices of these positions is pushed onto the stack. The stack must
therefore be increased in size before this happens. By pushing the last
array copied onto the stack, an entire set of cache misses is avoided.
Unfortunately, since a multi-partition is only done when the array is
more than twice the size of the cache, this reduction in misses in not
significant.

While emptying the containers back into the array, an opportunity is
taken\footnote{Results do not reflect this step. See \vref{bug1}.} to
find the smallest key in the leftmost partition. This is placed as a sentinel,
ensuring that every partition has a sentinel, and that no bounds checks are
required.

@@

The instruction count, shown in Figure \ref{quick instructions} is as predicted. The expected increases in instruction
count for multi-quicksort was due to a test with a smaller data set and smaller
constants. The actual results were not as high as the expected ones, but closer
to 20\% and 45\% greater, for sequential and binary search versions,
respectively.

The cache results in Figure \ref{quick level 2 misses} were similar to
predictions, except that the multi-quicksorts did not behave as expected, and
performed worse than the memory-tuned version. However, at the edge of the graph
their slopes are considerably less than the slopes of the other sorts, and may
result in lower misses for larger data sets. The tuned version performed better
than the base version, as predicted.

@@

The sequential search stops the increase of branch misses, since every key will
only take one miss to find its proper place. The results of this on cycle count,
shown in Figure \ref{quick cycles} are enough to make up for the increase in
instruction count and level 1 accesses, both of which were significantly reduced
by using the binary search. The tuned quicksort, however, had the lower running
time and even base quicksort out-ran the multi-quicksorts.

LaMarca's results are again similar to the results presented here, but not
identical. The shape of the instruction count graph is the same in both cases,
but that of the cache misses is not. The ratio of the cache misses between base
and cache quicksort are the same in both sets of results, but the number of
cache misses in multi-quicksort is much higher here than it is in LaMarca's
findings. As with mergesort though, while the results are different, the shape
is similar: multi-quicksort has a much flatter slope than the other sorts.  This
difference is likely to be due to the implementation of multi-quicksort; a
slightly different approach, which is described immediately below, may have the
effect of changing the results sufficiently that they may agree with LaMarca's.

\section{Future Work}
In multi-quicksort, it may be possible to reduce the cache misses by sorting
each containerful of keys as soon as it is put back into the array. This step would
significantly increase the temporal locality, and may reduce the number of cache
misses by half.

The threshold in quicksort is the point at which a simpler sort is used to sort
the partition. A larger threshold will results in fewer checks for a sentinel. A
smaller threshold means that the instruction count due to the insertion sort
will be lower, but more partition steps may be taken. A smaller threshold also
makes it more likely that the keys to be sorted all fit in a single cache
block. However, the partition is likely to occur across a cache block boundary,
though this is not necessarily a bad thing; the next partition to be sorted will
be adjacent to the current block, and will already be in memory. Observing how
results, especially instruction count and branch misses, vary with changes 
to the threshold may be interesting.

### 2005-06-14 docs/tech_report/docs/intro.tex
moved doc dirs into docs/ moved scripts into scripts/ moves counters/ into data/

Chapter \ref{background} provides background information on caches and branch predictors. It
also discusses the theory of sorting, and conventions used in the algorithms in
this project.

Chapter \ref{method}  discusses the method and framework used in this project, including the
tools used and the reasons for using them.

>>>>

Chapter \ref{background} provides background information on caches and branch predictors. It
also discusses the theory of sorting, and conventions used in the algorithms in
this report.

Chapter \ref{method} discusses the method and framework used in this research, including the
tools used and the reasons for using them.

### 2005-06-14 docs/tech_report/docs/quick.tex
moved doc dirs into docs/ moved scripts into scripts/ moves counters/ into data/

LaMarca uses the optimised quicksort described by Sedgewick for his base
quicksort. This is an iterative quicksort which uses an explicit stack. In the
worst case, this stack must be the same size as the array, and must be
initialised at the start of the algorithm. To reduce the size of the stack,
Sedgewick proposes sorting the smallest partition first. This optimisation
reduces the maximum size of the stack to $O(logN)$. In practice, two \cc{int}s,
the left and the right index, are required for each partition.

>>>>

LaMarca uses the optimized quicksort described by Sedgewick for his base
quicksort. This is an iterative quicksort which uses an explicit stack. In the
worst case, this stack must be the same size as the array, and must be
initialised at the start of the algorithm. To reduce the size of the stack,
Sedgewick proposes sorting the smallest partition first. This optimization
reduces the maximum size of the stack to $O(logN)$. In practice, two \cc{int}s,
the left and the right index, are required for each partition.
